---
title: "DSFieldFinal"
author: "Caden Zonnefeld"
date: "11/23/2022"
output: pdf_document
---
## Purpose and Nature of the Data
The question of interest is if COVID-19 equally effected the United States or if certain states/regions took a harder hit from the virus. Furthemore, how does the United States response compare to other countries around the world? I will examine trends and comparisons of the data to explore this question. 

The data used to address this question is sourced from John Hopkins University. They have a public GitHub repository that contains information about number of cases and deaths from the the COVID-19 virus at a national and global level. The data is stored in four separate .csv files corresponding to global cases, US cases, global deaths, and US cases.


## Data Collection
Reading the data in from the GitHub repository.
```{r data import, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)

url_begin <- 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/'

files <- c('time_series_covid19_confirmed_global.csv',
           'time_series_covid19_deaths_global.csv',
           'time_series_covid19_confirmed_US.csv',
           'time_series_covid19_deaths_US.csv')

urls <- str_c(url_begin, files)


global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
```



## Wrangling the Data
Pivoting the data into a format that is more friendly for future analysis.
End up with data that shows Province/State, Country, Date, Number of Cases, and Number of Deaths. Furthermore, a population feature is added to the global dataset.
```{r data wrangling, echo=TRUE, message=FALSE, warning=FALSE}
global_cases <- global_cases %>%
      pivot_longer(cols = -c('Province/State', 'Country/Region',
                             'Lat', 'Long'),
                   names_to = 'date',
                   values_to = 'cases') %>%
      select(-c('Lat', 'Long'))

global_deaths <- global_deaths %>%
      pivot_longer(cols = -c('Province/State', 'Country/Region',
                             'Lat', 'Long'),
                   names_to = 'date',
                   values_to = 'deaths') %>%
      select(-c('Lat', 'Long'))

global <- global_cases %>%
      full_join(global_deaths) %>%
      rename(Country_Region = 'Country/Region',
             Province_State = 'Province/State') %>%
      mutate(date = mdy(date))

US_cases <- US_cases %>%
      pivot_longer(cols = -(UID:Combined_Key),
                   names_to = 'date',
                   values_to = 'cases') %>%
      select(UID:cases) %>%
      mutate(date = mdy(date)) %>%
      select(-c(iso2, iso3, code3, FIPS, Lat, Long_))

US_deaths <- US_deaths %>%
      pivot_longer(cols = -(UID:Combined_Key),
                   names_to = 'date',
                   values_to = 'deaths') %>%
      select(UID:deaths) %>%
      mutate(date = mdy(date)) %>%
      select(-c(iso2, iso3, code3, FIPS, Lat, Long_))

US <- US_cases %>%
      full_join(US_deaths)

global <- global %>%
          unite('Combined_Key',
                c(Province_State, Country_Region),
                sep=',',
                na.rm = TRUE,
                remove = FALSE)

US_by_state <- US %>%
      group_by(Province_State, Country_Region, date) %>%
      summarize(cases = sum(cases), deaths = sum(deaths)) %>%
      select(Province_State, Country_Region, date, cases, deaths) %>%
      ungroup()

US_totals <- US_by_state %>%
      group_by(Country_Region, date) %>%
      summarize(cases = sum(cases), deaths = sum(deaths)) %>%
      select(Country_Region, date, cases, deaths) %>%
      ungroup()
```

## Initial Data Examination and Population Join

Viewing the basic summary statistics of the data and performing a gut check for the ranges of values.

```{r data examination, echo=TRUE, message=FALSE, warning=FALSE}
summary(global)

global <- global %>% filter(cases > 0)

summary(global)

global %>% filter (cases > 98000000)
```



## Visualizing the Cleaned Data

Showing the increase in cases and deaths for all of the United States then just Arizona and California in particular. The gradual increase over time makes sense but this covers over some of the smaller trends. I will investigate this by including some lagged features to see the microtrends that occur in the data.

```{r data viz, echo=TRUE, message=FALSE, warning=FALSE}

US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y = cases)) + 
  geom_line(aes(color = 'cases')) +
  geom_point(aes(color = 'cases')) +
  geom_line(aes(y=deaths, color = 'deaths')) +
  geom_point(aes(y=deaths, color = 'deaths')) +
  scale_y_log10() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = 'COVID19 in US', y = NULL)

US_by_state %>%
  filter(Province_State == 'Arizona') %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y = cases)) + 
  geom_line(aes(color = 'cases')) +
  geom_point(aes(color = 'cases')) +
  geom_line(aes(y=deaths, color = 'deaths')) +
  geom_point(aes(y=deaths, color = 'deaths')) +
  scale_y_log10() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = 'COVID19 in Arizona', y = NULL)

US_by_state %>%
  filter(Province_State %in% c('California')) %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y = cases)) + 
  geom_line(aes(color = 'cases')) +
  geom_point(aes(color = 'cases')) +
  geom_line(aes(y=deaths, color = 'deaths')) +
  geom_point(aes(y=deaths, color = 'deaths')) +
  scale_y_log10() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = 'COVID19 in California', y = NULL)

```

## Enriching the Dataset
Adding lagged features to examine the progression of cases and deaths due to COVID19. Also visualizing the progression of new cases/deaths. This view of the data gives a more insightful perspective as it better illustrates the recent developments of the COVID19 virus. Note that the increased variability toward the present is due to inconsistency in data collection/sporadic results. For example, we can conclude that Colorado's new cases peaked around the beginning of 2022. Next I examined West Coast states and noticed a similar trend among each of the stats for infections and deaths from the COVID19 virus. Following this I considerd New York and California, two of the most populous states in the country to see if and how they responded differently. New York suffers a greater initial spike upon the onset of the virus; however, the two states follow a similar trend folliwng that. The large spike likely is a result of high population density in New York. Finally, I was surprised to see that New York (a high population state) and Wyoming (a low population state) experienced similar trends in response to COVID19. My initial hypothesis was that lower population states would experience differing effects.

```{r feature engineering, echo=TRUE, message=FALSE, warning=FALSE}

US_by_state <- US_by_state %>%
        mutate(new_cases = cases - lag(cases),
               new_deaths = deaths - lag(deaths))

US_totals <- US_totals %>%
        mutate(new_cases = cases - lag(cases),
               new_deaths = deaths - lag(deaths))

global <- global %>%
        mutate(new_cases = cases - lag(cases),
               new_deaths = deaths - lag(deaths))

tail(US_totals %>% select(new_cases, new_deaths, everything()))

US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y = new_cases)) + 
  geom_line(aes(color = 'new_cases')) +
  geom_point(aes(color = 'new_cases')) +
  geom_line(aes(y=new_deaths, color = 'new_deaths')) +
  geom_point(aes(y=new_deaths, color = 'new_deaths')) +
  scale_y_log10() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = 'Rolling COVID19 in US', y = NULL)

US_by_state %>%
  filter(Province_State == 'Colorado') %>% 
  filter(cases > 0) %>%
  ggplot(aes(x=date, y = new_cases)) + 
  geom_point(aes(color = 'new_cases')) +
  geom_point(aes(y=new_deaths, color = 'new_deaths')) +
  scale_y_log10() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = 'Rolling COVID19 in Colorado', y = NULL)

US_by_state %>%
  filter(Province_State %in% c('Washington', 'Oregon', 'California')) %>% 
  filter(cases > 0) %>%
  ggplot(aes(x=date, y = new_cases)) + 
  geom_point(aes(color = Province_State)) +
  geom_point(aes(y=new_deaths, color = Province_State)) +
  scale_y_log10() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = 'Rolling COVID19 on the West Coast', y = NULL)

US_by_state %>%
  filter(Province_State %in% c('New York', 'California')) %>% 
  filter(cases > 0) %>%
  ggplot(aes(x=date, y = new_cases)) + 
  geom_point(aes(color = Province_State)) +
  geom_point(aes(y=new_deaths, color = Province_State)) +
  scale_y_log10() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = 'Rolling COVID19 New York v. California', y = NULL)


US_by_state %>%
  filter(Province_State %in% c('New York', 'Wyoming')) %>% 
  filter(cases > 0) %>%
  ggplot(aes(x=date, y = new_cases)) + 
  geom_point(aes(color = Province_State)) +
  geom_point(aes(y=new_deaths, color = Province_State)) +
  scale_y_log10() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = 'Rolling COVID19 New York v. Wyoming', y = NULL)
```

## Evaluting the United States Response
This bar chart shows that the the effectiveness of the United States response to the COVID19 virus paled in comparison to that of the other most populous countries in the world. Even though China and India have vastly larger populations, they experienced far less cases. Focusing in on a comparison between the United States and India (a country with 3x the population), we can see that India has done a better job of prevention as their infection and fatality rate are decreasing while the United States rates have stayed relatively the same.
```{r evaluating US Response, echo=TRUE, message=FALSE, warning=FALSE}
neighbors <- c('Pakistan', 'Indonesia', 'US', 'India', 'China')

global %>%
  filter(Country_Region %in% neighbors) %>%
  group_by(Country_Region) %>%
  summarize(cases = sum(cases)) %>%
  ggplot(aes(x=Country_Region)) +
  geom_col(aes(y=cases, fill = Country_Region)) +
  labs(title = 'COVID19 Cases of 5 Most Populous Countries')

global %>%
  filter(Country_Region %in% neighbors) %>%
  group_by(Country_Region) %>%
  summarize(deaths = sum(deaths)) %>%
  ggplot(aes(x=Country_Region)) +
  geom_col(aes(y=deaths, fill = Country_Region)) +
  labs(title = 'COVID19 Deaths of 5 Most Populous Countries')
 

global %>%
  filter(Country_Region %in% c('US', 'India')) %>% 
  filter(cases > 0) %>%
  ggplot(aes(x=date, y = new_cases)) + 
  geom_point(aes(color = Country_Region)) +
  geom_point(aes(y=new_deaths, color = Country_Region)) +
  scale_y_log10() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = 'Rolling COVID19 US v. India', y = NULL)
```

## Modeling Data

Comparing the two least populous states based on COVID19 cases/deaths. Though Wyoming experienced a statistically significant increase in cases they did not have a statistically significant difference in deaths. Furthermore, we can notice that the US had a statistically significant amount greater cases and deaths than the much larger nation of India. 

```{r modeling, echo=TRUE, message=FALSE, warning=FALSE}
smallest <- US_by_state %>%
        filter(Province_State %in% c('Vermont', 'Wyoming'))
linear <- lm(cases ~ Province_State, data = smallest)
summary(linear)

linear <- lm(deaths ~ Province_State, data = smallest)
summary(linear)


neighbor_countries <- global %>%
        filter(Country_Region %in% c('US', 'India'))
linear <- lm(cases ~ Country_Region, data = neighbor_countries)
summary(linear)

neighbor_countries <- global %>%
        filter(Country_Region %in% c('US', 'India'))
linear <- lm(deaths ~ Country_Region, data = neighbor_countries)
summary(linear)
```

## Biases and Conclusion
This dataset had several areas in which bias may have affected the results. For instance, the reporting practices of COVID19 cases differences greatly across different countries and even regions within a country. Furthermore, more desolate regions of countries like India or China may be under-counted or even flat out ignored. Another place that bias may have creeped into this data is by using raw counts instead of rates by incorporating populations. Unfortunately, my computer was unable to access the population .csv file that was used in the lectures.

Despite some anomalies, my conclusion is that the COVID19 pandemic affected the United States in a similar fashion across the board in regards to deaths, though different regions produced differing numbers of cases. On the other hand, preliminary analysis and visualization indicates that the United States did a poor job handling the pandemic compared to other populous countries in the world. This is clearly evident by vastly greater counts of cases and deaths and was confirmed with a linear model that compared to India, though the same could be done for any other peer country with similar populations. 



 
      
