---
title: "NYPD Shooting Incident Data Report"
author: "Caden Zonnefeld"
date: "10/26/2022"
output: pdf_document
---

## Data Collection

```{r data import, message=FALSE}
library(tidyverse)
library(lubridate)

url <- 'https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD'
nypd <- read_csv(url, show_col_types = FALSE)

```

Collects data pertaining to NYPD shooting incidents from data.gov.
The data is accessed by utilizing a URL and reading the data in via a .csv file.
The raw data consists of nearly 26k observations and 19 features. The data
acquisition is reproducible since it is collected from a pubplic URL.

## Tidying the Data
```{r tidy and transform data}

nypd$OCCUR_DATE <- mdy(nypd$OCCUR_DATE)

analysis_vars <- c('OCCUR_DATE', 'OCCUR_TIME', 'BORO', 'STATISTICAL_MURDER_FLAG',
                   'VIC_AGE_GROUP', 'VIC_SEX', 'VIC_RACE')
nypd <- nypd[analysis_vars]

nypd

```

Inspecting the data and casting the OCCUR_DATE field to a date object to be
properly treated as a date. Furthermore, reducing the dataset to the features
I will use in an analysis that will focus on the victims of such incidents.
The data is now tidy and ready for further analysis.

## Data Visualization and Analysis

```{r visualizing data, victim sex}
boro_numbers <- count(nypd %>% group_by(BORO))

boro_numbers$male <-count(nypd %>% group_by(BORO) %>% filter(VIC_SEX == 'M'))$n
boro_numbers$female <-count(nypd %>% group_by(BORO) %>% filter(VIC_SEX == 'F'))$n

boro_numbers %>% mutate(male_prop = male/n, female_prop = female/n)
```

The table indicates the the victim count and proportion of victims by gender
for each of New York's boroughs. The data reveals that there are far more males
that have been involved in incidents than females.

```{r visualizing data, borough}
ggplot(nypd) + geom_bar(aes(y=BORO, fill = BORO)) + 
  ggtitle('Incidents by Borough') + xlab('Count') + ylab('New York Borough')
```
The first bar graph illustrates the number of incidents that occurred in each
borough. We can gather that Staten Island has far fewer incidents while
Brooklyn has the most recorded incidents. In fact, Brooklyn appears to have 
about 10x as many incidents as Staten Island.

```{r visualizing data, race}
ggplot(nypd) + geom_bar(aes(y=VIC_RACE, fill = VIC_RACE)) +
  ggtitle('Incidents by Victim Race') + xlab('Count') + ylab('Race')
```
The next bar graph breaks down the race of the victim in each incident. Note
that black and to a lesser extent Hispanic individuals account for far more
incidents than other races. It appears that black individuals are cited in
incidents at disproportionate rates to their portion of the population.

```{r visualizing data, age group}
nypd_date <- nypd %>% filter(VIC_AGE_GROUP != 'UNKNOWN') %>% 
  group_by(YEAR = year(OCCUR_DATE)) %>% count(VIC_AGE_GROUP)
nypd_date

minors <- nypd_date %>% filter(VIC_AGE_GROUP == '<18')


minors %>% ggplot(aes(x=YEAR, y=n)) + geom_line() + ylab('Count') + xlab('Year') +
  ggtitle('Number of Incidents Involving Minors by Year')

```
The grouped table shows the number of incidents that involved people from each
age group band in the specified year. Lastly, the line graph shows the number of
incidents involving minors on a yearly basis. Incidents involving minors have
been greatly reduced since the beginning of data collection; however, there have
been an uptick in such incidents over the last few years.


```{r analyzing data, age group lm}
age_model <- lm(n ~ VIC_AGE_GROUP, data = nypd_date)
summary(age_model)
```
A simple linear model used to predict the number of shooting incidents in a
given year by using only victim age group band indicates that age group 65+ has
a significant effect that lowers the expected number of incidents. Conversely,
age groups 18-24 and 25-44 have significant effects that increase the expected
number of shooting incidents. Interestingly, victim age group alone accounts
for 84% of the variation in the data.

```{r analyzing data, }
nypd_race <- nypd %>% filter(VIC_AGE_GROUP != 'UNKNOWN') %>% 
  group_by(YEAR = year(OCCUR_DATE)) %>% count(VIC_RACE)

race_model <- lm(n ~ VIC_RACE, data = nypd_race)
summary(race_model)
```
Once again I used a simple linear regression to predict the number of shooting
incidents in a given year using victim race as a predictor. We can see that a 
race of Hispanic or black have a significant positive effect on the predicted
number of incidents involved in for a given year. This model has good
explanatory power since it accounts for 91% of the variation in the data.


## Bias Recognition and Conclusion
For full disclosure, I am a white male which may have implicitly biased my
approach to considering the data, particularly the race and gender features. 

The data indicates that non-white individuals, particularly those who are black,
are at a far greater risk of being victimized by a shooting incident. The stark
contrast makes me wonder what biases against black individuals put them at such
an elevated risk to be involved in a shooting incident.
Furthermore, we can see that men are more prone to being victimized as there is 
about an 8:1 ratio for men to women victimized buy shootings. One encouraging
trend from the data is the decrease of minors being involved in shootings;
however, there is some cause for concern as the number of victimized minors has
increased each year since 2019. Though we do not have the population of each 
borough, we can see that by number of incidents alone Staten Island appears to
be the safest as it has less than a third of the incidents of the next lowest 
borough.